---
title: "What affects the populartiy of horror movies"
date: "2026-05-01"
---

# What affects the popularity of horror movies

### Introduction to the Dataset

The primary objective of this project is to investigate the factors that influence the popularity of horror movies. The analysis is based on the Horror Movies dataset provided by the TidyTuesday, which source is from The Movie Database (TMDB).

The raw dataset contains 32,540 records and 20 variables, documenting the evolution of the horror genre from the 1950s to the present. To provide a structured analysis, we have categorized the most significant variables as follows:

-   Identification:

    -   title: The movie's title (original or translated into English).

    -   original_language: The primary language, indicating the movie's country of origin.

    -   release_date: The official date the movie was made available to the public.

-   Popularity & Ratings:

    -   popularity: A metric calculated by TMDB’s internal algorithm.

    -   vote_count: The total number of user ratings received.

    -   vote_average: The mean score (out of 10) provided by the audience.

-   Financial Metrics:

    -   budget: The total capital spent on the film's production.

    -   revenue: Earnings generated by the movie.

-   Technical and Structural Aspects:

    -   runtime: The duration of the film in minutes.

    -   collection_name: Indicates if the movie is part of a larger franchise or series.

In this study, popularity serves as the dependent variable. Analysis aims to determine how financial investment, technical characteristics, and audience reception correlate with a film’s overall popularity in the market.

# Libraries

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

library(lubridate)
library(stringr)
```

# The dataset

### Download the dataset

```{r}
df <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-11-01/horror_movies.csv')

```

# Data cleaning and Preparation

### Variable selection

```{r}
df %>%
  colnames()
```

```{r}
df %>%
  sapply(class) %>%
  as.data.frame()
```

Checking the difference between title and original_title

```{r}
df %>%
  select(original_title, title) %>%
  tail(10)
```

To improve readability, only title will be used.

Checking usability of particular variables

```{r}
df %>%
  select(overview) %>%
  head(5)
```

```{r}
df %>%
  select(tagline) %>%
  head(5)
```

Overview or tagline describes the movie which won't be useful in the analysis.

```{r}
df %>%
  select(genre_names) %>%
  head(5)
```

The genre_names variable contains multiple categories for a single movie. It will be handled in the EDA phase by creating new columns.

```{r}
df %>%
  select(collection) %>%
  head(5)
```

Name of the collection won't be useful.

```{r}
df %>%
  select(collection_name) %>%
  head(10)
```

```{r}
df %>%
  select(adult) %>%
  head(5)
```

```{r}
df %>%
  select(backdrop_path) %>%
  head(5)
```

```{r}
df %>%
  ggplot(aes(x = status)) +
  geom_bar()
```

The status variable consists almost entirely of the "Released" category. Since it lacks variance, it does not provide any predictive power and is redundant.

### Removing unnecessary variables

```{r}
df <- df %>%
  subset(select= -c(original_title, id, overview, tagline, poster_path, collection, backdrop_path, status ) )
```

```{r}

df %>%
  colnames() %>%
  as.data.frame()
```

Ended up with 10 predictors.

### Correcting errors

Remove NAs where it is neccessary

```{r}
df <- df %>%
  drop_na(original_language, release_date, popularity,vote_count, vote_average, budget, revenue, runtime)

```

Check correction of numerical data

\*For better readability

```{r}
options(scipen = 999)
```

```{r}

df %>%
  summarise(
    across(
      c(popularity, vote_count, vote_average, budget, revenue, runtime),
      list(
        avg = ~mean(.x),
        median = ~median(.x),
        max = ~max(.x),
        min = ~min(.x)
      )
    )
  ) %>%
  pivot_longer(
    everything(),
    names_to = "Variable",
    values_to = "Value"
  )
```

Visualization of suspicious variables

```{r}
df %>%
  ggplot(aes(x = popularity)) +
  geom_histogram(bins = 50 ) 
```

```{r}
df %>%
  select(title, popularity) %>%
  arrange(desc(popularity)) 
```

Verified the authenticity of the outlier. The exceptionally high popularity score for "Orphan: First Kill" was confirmed to be legitimate based on TMDB trends at the time, so it remains in the dataset.

Votes are variable only if more than 50 people have voted

```{r}
df <- df %>%
  filter(vote_count >20, runtime > 0, budget > 0, revenue > 0)
```

The decision to filter movies with more than 50 votes was made to ensure statistical significance of the range variable.

Excluded records with 0 values in budget and revenue, as they likely represent missing data.

Check if languages are coherent

```{r}
unique(df$original_language)
```

Remaining observations

```{r}
df %>%
  count()
```

While the original dataset contained over 32.000 records, by filtering for movies with confirmed financial data and at least 20 audience votes, the vast majority has to be removed. Due to that, the analysis will be far more reliable.

# Explanatory Data Analysis

```{r}
as.data.frame(glimpse(df)) 
```

### Transform release_date column into year and month columns

```{r}
df <- df %>%
  mutate(
    year_added = year(release_date),
    month_added = month(release_date)
  )
```

In order to check if in current years the popularity of horror movies was higher then usual and etc with months (for example if halloween affects the popularity)

### Diversion of original languages

```{r}
df %>%
  select(original_language) %>%
  count(original_language, sort = TRUE) 
  
  
```

Diversion 5 the most common appeard countires and rest into others column.

```{r}
df <- df %>%
  mutate(
    en_language = if_else(original_language == "en", 1, 0),
    ja_language = if_else(original_language == 'ja', 1, 0),
    es_language = if_else(original_language == 'es', 1, 0),
    ko_language = if_else(original_language == 'ko', 1, 0),
    fr_language = if_else(original_language == 'fr', 1, 0),
    
    others_language = if_else(!(original_language %in% c('en', 'ja', 'es', 'ko', 'fr')), 1,0)
    
         )
```

### Change collection name to is_in_collection

```{r}
df <- df %>%
  mutate(
    is_in_collection = if_else(!is.na(collection_name), 1,0)
  )
```

### Genre of movies into columns

```{r}
df%>%
  select(genre_names) %>%
  distinct() 
```

A few of the most important genre into columns

```{r}
df <- df %>%
  mutate(
    genre_horror = as.numeric(str_detect(genre_names, "Horror")),
    genre_thriller = as.numeric(str_detect(genre_names, "Thriller")),
    genre_fantasy = as.numeric(str_detect(genre_names, "Fantasy")),
    genre_sc = as.numeric(str_detect(genre_names, "Science Fiction")),
    genre_crime = as.numeric(str_detect(genre_names, "Crime")),
    genre_action = as.numeric(str_detect(genre_names, "Action")),
    genre_comedy = as.numeric(str_detect(genre_names, "Comedy")),
    genre_drama = as.numeric(str_detect(genre_names, "Drama"))
  ) 
```

### Profit column

Revenue - budget - did the film earn or lost money

```{r}
df <- df %>%
  mutate(
    profit = revenue - budget
  )
```

### New schema of columns

```{r}
colnames(df)
```

### Do years or months impact the popularity of horror movies?

```{r}
df %>%
  ggplot(aes(x = year_added, y = popularity)) +
  geom_point() +
  scale_y_log10() +
  labs(title = "Horror Movie Popularity Over the Years",
       x = "Year of Release",
       y = "Popularity ")
    

```

To investigate if there is a long-term trend in the popularity of horror movies.

A scatter plot shows a significant rise of popularity in recent years.

A crucial bias is the fact that popularity metric is dynamic and is favouring recent releases.

```{r}
df %>%
  ggplot(aes(x = month_added, y = popularity)) +
  geom_point() +
  scale_y_log10() +
  labs(title = "Monthly Trends in Horror Movie Popularity",
       x = "Month ",
       y = "Popularity")
```

To see if the timing of a movie's release (e.g., around Halloween) impacts its overall popularity score.

The scores are relatively spread, but with slight dips at the very beginning and end of the year.

### Does the language of the movie affects its success?

```{r}
df %>%
  filter(original_language %in% c("en", "ja", "es", "ko", "fr")) %>%
  ggplot(aes(x = original_language, y = popularity, fill = original_language)) +
  geom_boxplot() +
  scale_y_log10()  +
  labs(title = "Popularity Distribution by Movie Language",
       x = "Original Language",
       y = "Popularity")

```

To compare the market reach of English-language horror movies against major international competitors like Japanese or Korean horror.

While English has the most data points, Korean (ko) and Japanese (ja) horror films show very high median popularity levels.

### Amount of specific genres

```{r}

genre_sums <- data.frame(
  Genre = c("Horror", "Action", "Thriller", "Fantasy", "Comedy", "Science Fiction", "Crime", "Drama"),
  Amount = c(sum(df$genre_horror), sum(df$genre_action), sum(df$genre_thriller), 
             sum(df$genre_fantasy), sum(df$genre_comedy), sum(df$genre_sc), sum(df$genre_crime), sum(df$genre_drama))
)

ggplot(genre_sums, aes(x = Genre, y = Amount, fill = Genre)) +
  geom_col() +
  labs(title = "The amount of movies in genres",
       x = "Genre",
       y = "Number of Movies")
```

To identify which genres are most commonly paired with horror in the industry.

### The aspect of money in popularity

```{r}
df %>%
  pivot_longer(
    cols = c(profit, revenue, budget), 
    names_to = "money_type", 
    values_to = "amount") %>%
  ggplot(aes(x = amount, y = popularity, color = money_type)) +
  geom_point() +
  facet_wrap(~money_type) +
  scale_y_log10()  +
  labs(title = "Impact of Financial Metrics on Popularity",
       x = "Amount",
       y = "Popularity")
```

To see if spending more money (Budget) or earning more (Revenue/Profit) directly correlates with higher popularity.

Revenue and Profit show a clearer positive correlation with popularity than the initial Budget does.

### Impact of being part of a collection

```{r}
df %>%
  mutate(is_in_collection = if_else(is_in_collection == 1, "Franchise", "Standalone")) %>%
  ggplot(aes(x = is_in_collection, y = popularity, fill = is_in_collection)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(title = "Franchise vs. Standalone Horror Movies",
       x = "Movie Category",
       y = "Popularity ")
```

To test if movies that are part of a collections have an advantage in popularity over one-off films.

The "Franchise" group has a noticeably higher median popularity score.

### Audience engagement

```{r}
df %>%
ggplot( aes(x = vote_count, y = popularity)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Audience Engagement: Vote Count vs. Popularity",
       x = "Number of Votes ",
       y = "Popularity")
```

To verify if the popularity score is a reflection of total audience interaction.

The plot reveals a strong positive correlation between the number of votes and popularity. There are also undefined.

### Distribution of movie lengths

```{r}
df %>%
  ggplot( aes(x = runtime)) +
  geom_histogram() +
  labs(title = "Distribution of Horror Movie Runtimes",
       x = "Runtime (minutes)",
       y = "Frequency")
```

To identify the "industry standard" for horror movie durations and to spot any unusual outliers (extremely short or long films).

A histogram showing a strong peak between 80 and 100 minutes.

### Are the highest ranked the most popular?

```{r}
df %>%
  ggplot( aes(x = vote_average, y = popularity)) +
  geom_point() +
  scale_y_log10() +
  labs(title = "Quality vs. Fame: Rating vs. Popularity",
       x = "Average Vote",
       y = "Popularity")
```

To see if higher rated movies lead to higher popularity.

There is a slight upward trend.
