---
title: "What affects the populartiy of horror movies"
authors: Aleksandra Duda, Karolina Jajko, Yehor Kharchenko, Wiktoria Wachel
date: "2026-05-01"
---

# What affects the popularity of horror movies

### Introduction to the Dataset

The primary objective of this project is to investigate the factors that influence the popularity of horror movies. The analysis is based on the Horror Movies dataset provided by the TidyTuesday, which source is from The Movie Database (TMDB).

The raw dataset contains 32,540 records and 20 variables, documenting the evolution of the horror genre from the 1950s to the present. To provide a structured analysis, we have categorized the most significant variables as follows:

-   Identification:

    -   title: The movie's title (original or translated into English).

    -   original_language: The primary language, indicating the movie's country of origin.

    -   release_date: The official date the movie was made available to the public.

-   Popularity & Ratings:

    -   popularity: A metric calculated by TMDB’s internal algorithm.

    -   vote_count: The total number of user ratings received.

    -   vote_average: The mean score (out of 10) provided by the audience.

-   Financial Metrics:

    -   budget: The total capital spent on the film's production.

    -   revenue: Earnings generated by the movie.

-   Technical and Structural Aspects:

    -   runtime: The duration of the film in minutes.

    -   collection_name: Indicates if the movie is part of a larger franchise or series.

In this study, popularity serves as the dependent variable. Analysis aims to determine how financial investment, technical characteristics, and audience reception correlate with a film’s overall popularity in the market.

# Libraries

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

# The dataset

### Download the dataset

```{r}
df <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-11-01/horror_movies.csv')

```

# Data cleaning and Preparation

### Variable selection

```{r}
df %>%
  colnames()
```

```{r}
df %>%
  sapply(class) %>%
  as.data.frame()
```

Checking the difference between title and original_title

```{r}
df %>%
  select(original_title, title) %>%
  tail(10)
```

To improve readability, only title will be used.

Checking usability of particular variables

```{r}
df %>%
  select(overview) %>%
  head(5)
```

```{r}
df %>%
  select(tagline) %>%
  head(5)
```

Overview or tagline describes the movie which won't be useful in the analysis.

```{r}
df %>%
  select(genre_names) %>%
  head(5)
```

The genre_names variable contains multiple categories for a single movie. It will be handled in the EDA phase by creating new columns.

```{r}
df %>%
  select(collection) %>%
  head(5)
```

Name of the collection won't be useful.

```{r}
df %>%
  select(collection_name) %>%
  head(10)
```

```{r}
df %>%
  select(adult) %>%
  head(5)
```

```{r}
df %>%
  select(backdrop_path) %>%
  head(5)
```

```{r}
df %>%
  ggplot(aes(x = status)) +
  geom_bar()
```

The status variable consists almost entirely of the "Released" category. Since it lacks variance, it does not provide any predictive power and is redundant.

### Removing unnecessary variables

```{r}
df <- df %>%
  subset(select= -c(original_title, id, overview, tagline, poster_path, collection, backdrop_path, status ) )
```

```{r}

df %>%
  colnames() %>%
  as.data.frame()
```

Ended up with 10 predictors.

### Correcting errors

Remove NAs where it is neccessary

```{r}
df <- df %>%
  drop_na(original_language, release_date, popularity,vote_count, vote_average, budget, revenue, runtime)

```

Check correction of numerical data

\*For better readability

```{r}
options(scipen = 999)
```

```{r}

df %>%
  summarise(
    across(
      c(popularity, vote_count, vote_average, budget, revenue, runtime),
      list(
        avg = ~mean(.x),
        median = ~median(.x),
        max = ~max(.x),
        min = ~min(.x)
      )
    )
  ) %>%
  pivot_longer(
    everything(),
    names_to = "Variable",
    values_to = "Value"
  )
```

Visualization of suspicious variables

```{r}
df %>%
  ggplot(aes(x = popularity)) +
  geom_histogram(bins = 50 ) 
```

```{r}
df %>%
  select(title, popularity) %>%
  arrange(desc(popularity)) 
```

Verified the authenticity of the outlier. The exceptionally high popularity score for "Orphan: First Kill" was confirmed to be legitimate based on TMDB trends at the time, so it remains in the dataset.

Votes are variable only if more than 50 people have voted

```{r}
df <- df %>%
  filter(vote_count >20, runtime > 0, budget > 0, revenue > 0)
```

The decision to filter movies with more than 50 votes was made to ensure statistical significance of the range variable.

Excluded records with 0 values in budget and revenue, as they likely represent missing data.

Check if languages are coherent

```{r}
unique(df$original_language)
```

Remaining observations

```{r}
df %>%
  count()
```

While the original dataset contained over 32.000 records, by filtering for movies with confirmed financial data and at least 50 audience votes, the vast majority has to be removed. Due to that, the analysis will be far more reliable.

# Explanatory Data Analysis

```{r}
as.data.frame(glimpse(df)) 
```

### Transform release_date column into year and month columns

```{r}
df %>%
  mutate(
    year_added = year(release_date)
  )
```
